---
title: "ATP matches analysis"
author: "Mateusz Majak and Przemysław Mazurek"
date: "12.09.2020"
output: 
  html_document: 
    number_sections: true
    fig_caption: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# About the dataset
The dataset used in this project consists of information about over 50 thousand ATP matches from years 2000-2017. It was created by Jeff Sackmann. The dataset was uploaded from kaggle.com and is available [here](https://www.kaggle.com/gmadevs/atp-matches-dataset). For more data and information about licence visit author's [github profile](https://github.com/JeffSackmann/tennis_atp).   

# Preparations {.tabset .tabset-fade .tabset-pills}

## Load libraries
We load multiple libraries for data manipulation and visualisation together with 
tools for data modelling.

```{r message = FALSE}

# data manipulation
library(countrycode)
library(extdplyr)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(janitor)

# Interactivity
library(plotly)
library(crosstalk)
library(DT)

#Shiny App
library(shiny)
library(stargazer)

#visualisations
library(corrplot)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(ggthemes)

#data modelling
library(bestNormalize)
```

## Functions
We load our shiny app together with functions for data processing and visualization.
```{r message=FALSE, warning=FALSE}
#Shiny app
source("app.r")

#Function used to extract the data chosen by the user with the optional Box Cox transformation of the dependent variable
source("var_selection.r")

#Function used to extract details of models to compare them, created by dr hab. Piotr Wójcik.
source("F_regression_metrics.r")

#Function extracting letters from string from the right side
StringRightCut <- function(string, num){
  substr(string, nchar(string)-num+1, nchar(string))
}

#Function extracting letters from string from the left side
StringLeftCut <- function(string, num){
  substr(string, 1, num)
}
```

## Load data
```{r results=FALSE, class.source = "fold-show"}
files <- list.files(pattern="*.csv")

atp <- files %>%
  setNames(nm = .) %>% 
  purrr::map_dfr(
    ~ read_csv(.x, col_types = cols(), col_names = FALSE),
    .id = "file_name"
  ) 
```

# Dataset preparation

At first, we add **Year** column based on file names.
```{r class.source = "fold-show"}
atp$file_name <- StringRightCut(atp$file_name,8)
atp$file_name <- StringLeftCut(atp$file_name,4)
atp[1,1] <- "Year"
```

And now we choose useful variables and convert them into doubles and factors.
This is how dataset look like after transformations:
```{r class.source = "fold-hide"}
atp <- 
  atp %>%
  janitor::row_to_names(1) %>%
  dplyr::select(Year, loser_ioc, winner_ioc, winner_name, loser_name, surface, tourney_level, winner_hand, winner_ht, winner_age, winner_rank, winner_rank_points, loser_hand, loser_ht, loser_age, loser_rank,
                loser_rank_points, minutes, w_ace, w_df, w_svpt, w_1stIn, w_1stWon, w_2ndWon, w_SvGms, w_bpSaved, w_bpFaced, l_ace, l_df, l_svpt,
                l_1stIn, l_1stWon, l_2ndWon, l_SvGms, l_bpSaved, l_bpFaced) %>%
  drop_na() %>%
  dplyr::filter(minutes >= 18, #shortest atp tennis match lasted 18 minutes so any time shorter than that is not possible and must be a mistake
                minutes <= 665, #longest documented tennis match lasted 11:05h, so this is an upper bound
                loser_hand != "U") #because of only 1 observation left in this category
  
double_col_names <- c(1,9:12,14:36)
factor_col_names <- c(2:8,13)
atp[,double_col_names] <- lapply(atp[,double_col_names], as.double)
atp[,factor_col_names] <- lapply(atp[,factor_col_names], as.factor)

atp <- atp %>%
  dplyr::filter_if(is.factor, all_vars(.!="")) %>% #this is the way NA's were coded for factors
  dplyr::mutate(., surface = droplevels(surface), winner_hand = droplevels(winner_hand), loser_hand = droplevels(loser_hand))

glimpse(atp)
```
# EDA

```{r warning=FALSE}
atp_surface_matches <- atp %>% extdplyr::pct_routine(Year, surface)
colnames(atp_surface_matches)[3] <- "PercentOfMatches"

atp_hand_won_matches <- atp %>% count(Year, winner_hand)
colnames(atp_hand_won_matches)[3] <- "WinNumber"

atp_hand_lost_matches <- atp %>% count(Year, loser_hand)
colnames(atp_hand_lost_matches)[3] <- "LoseNumber"

atp_hand_matches <-
  merge(atp_hand_won_matches, atp_hand_lost_matches, 
        by.x = c("Year","winner_hand"),
        by.y = c("Year","loser_hand") )

colnames(atp_hand_matches)[2] <- "Hand"

atp_hand_matches$winRatio <- 
  atp_hand_matches$WinNumber/(atp_hand_matches$WinNumber+atp_hand_matches$LoseNumber)
```

## Percent of wins over time by hand
```{r warning=FALSE, message=FALSE, include=FALSE}
atp_hand_won_plotly <- highlight_key(atp_hand_matches %>% 
                                              filter(Year<2015))

pal <- palette(brewer.pal(9, "Set3"))

gg <- atp_hand_won_plotly %>% 
  ggplot(aes(Year, winRatio, col = Hand, group = Hand)) +
  geom_line() +
  geom_point(size=2)
  labs(x = "Year", y = "Hand") +
  theme_tufte() +
  NULL
```

```{r warning=FALSE}
filter <- bscols(filter_select("Hand","Percent Of wins over time (Select a Hand):",
                               atp_hand_won_plotly, ~Hand, 
                               multiple = TRUE),
                 ggplotly(gg, dynamicTicks = TRUE),
                 widths = c(15, 15)
                 )
bscols(filter)
```

## Percent of matches by surface
```{r warning=FALSE, message=FALSE, include=FALSE}
atp_surface_matches_plotly <- highlight_key(atp_surface_matches %>% 
                                              filter(Year<2015))

pal <- palette(brewer.pal(9, "Set3"))

gg <- atp_surface_matches_plotly %>% 
  ggplot(aes(Year, PercentOfMatches, col = surface, group = surface)) +
  geom_line() +
  geom_point(size=2)
  labs(x = "Year", y = "Surface") +
  theme_tufte() +
  NULL
```

```{r warning=FALSE}
filter <- bscols(filter_select("Surfaces","Percent Of Matches over time (Select a Surface):", atp_surface_matches_plotly, ~surface, 
                               multiple = TRUE),
                 ggplotly(gg, dynamicTicks = TRUE),
                 widths = c(15, 15)
                 )

bscols(filter)
```

## countries Details {.tabset .tabset-fade .tabset-pills}
```{r}
atp_country_won_matches <- atp %>% count(Year, winner_ioc)
colnames(atp_country_won_matches)[2] <- "Country"
colnames(atp_country_won_matches)[3] <- "WinNumber"

atp_country_lost_matches <- atp %>% count(Year, loser_ioc)
colnames(atp_country_lost_matches)[2] <- "Country"
colnames(atp_country_lost_matches)[3] <- "LoseNumber"

atp$winner_country <- countrycode(atp$winner_ioc, "ioc", "country.name")
atp$winner_country <- as.factor(atp$winner_country)

atp$loser_country <- countrycode(atp$loser_ioc, "ioc", "country.name")
atp$loser_country <- as.factor(atp$winner_country)

country_codes_win <- atp_country_won_matches$Country
atp_country_won_matches$Country <- countrycode(country_codes_win, "ioc", "country.name") # "ioc" - International Olympic Committee

country_codes_lose <- atp_country_lost_matches$Country
atp_country_lost_matches$Country <- countrycode(country_codes_lose, "ioc", "country.name")
```

### Average winner rank points
```{r fig.width=10, fig.height=5 ,warning=FALSE, message=FALSE}
options(scipen=999) #to prevent scientific notation
rank_data = atp %>% 
  select(winner_name, winner_country, winner_rank_points) %>%
  group_by(winner_country) %>% 
  summarise(Count = n(),
            Avg_rank_points = mean(winner_rank_points))
                           
datatable(rank_data %>%
    select(winner_country, Avg_rank_points)%>%                       
    arrange(desc(Avg_rank_points)), 
    options = list(pageLength = 5))   


worldmap = map_data("world")
merged_data <- merge(x = worldmap, y = atp_country_won_matches, by.x = "region", by.y = "Country", all.x = TRUE) %>% arrange(order)
ggplot(data = merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = WinNumber)) +
  scale_fill_viridis() +
  labs(fill='Number of won matches')
```

### Average loser rank points
```{r fig.width=10, fig.height=5 ,warning=FALSE, message=FALSE}
options(scipen=999) #to prevent scientific notation
rank_data = atp %>% 
  select(loser_name, loser_country, loser_rank_points) %>%
  group_by(loser_country) %>% 
  summarise(Count = n(),
            Avg_rank_points = mean(loser_rank_points))
                           
datatable(rank_data %>%
    select(loser_country, Avg_rank_points)%>%                       
    arrange(desc(Avg_rank_points)), 
    options = list(pageLength = 5)) 


worldmap = map_data("world")
merged_data <- merge(x = worldmap, y = atp_country_lost_matches, by.x = "region", by.y = "Country", all.x = TRUE) %>% arrange(order)
ggplot(data = merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = LoseNumber)) +
  scale_fill_viridis() +
  labs(fill='Number of lost matches')
```

## Histogram
```{r echo = FALSE}
atp_numeric_vars <- 
  sapply(atp, is.numeric) %>% 
  which() %>%
  names()

inputPanel(
  selectInput("numeric_variable", label = "Variable:",
              choices = atp_numeric_vars, selected = "minutes"),
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(5, 10, 20, 35), selected = 20),
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)
```

```{r echo = FALSE}
renderPlot({
  hist(atp[,input$numeric_variable][[1]], probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "", main = "Numeric variables histogram")
  
  dens <- density(atp$winner_age, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Factors violin plot {.tabset .tabset-fade .tabset-pills}

```{r echo = FALSE}
inputPanel(
  selectInput("factor_variable", label = "Factor variable:",
              choices = c("surface","tourney_level","winner_hand","loser_hand"), 
              selected = "surface"))
```

### Factor variables vs Winner Age
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=winner_age)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.55) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Winner age")
})
```

### Factor variables vs Loser Age
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=loser_age)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Loser age")
})
```

### Factor variables vs Winner height
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=winner_ht)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Winner height")
})
```

### Factor variables vs Loser height
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=loser_ht)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Loser height")
})
```

### Factor variables vs Winner rank
```{r echo = FALSE}

  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=winner_rank)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Winner rank")
})
```

### Factor variables vs Loser rank
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=loser_rank)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Loser rank")
})
```

### Factor variables vs Winner rank points
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=winner_rank_points)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Winner rank points")
})
```

### Factor variables vs Loser rank points
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=loser_rank_points)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Loser rank points")
})
```

### Factor variables vs Minutes
```{r echo = FALSE}
  renderPlot({
  ggplot(data=atp, aes(x=atp[,input$factor_variable][[1]], 
                       y=minutes)) +
    geom_violin(aes(fill = atp[,input$factor_variable][[1]]), width=1, size=0.2, 
                alpha=0.4) +
    theme(legend.position="none") +
    coord_flip() +
    xlab(input$factor_variable) +
    ylab("Minutes")
})
```

## Relationship between numeric variables
```{r fig.width=5, fig.height=4}
inputPanel(
  selectInput("x_axis", label = "x axis:",
              choices = c("minutes","winner_ht","winner_age","winner_rank","winner_rank_points",
                          "w_ace","w_df","w_svpt","w_1stIn","w_1stWon",
                          "w_2ndWon","w_SvGms","w_bpSaved","w_bpFaced"), 
              selected = "winner_ht"),
  selectInput("y_axis", label = "y axis:",
              choices = c("minutes","loser_ht","loser_age","loser_rank","loser_rank_points",
                          "l_ace","l_df","l_svpt","l_1stIn","l_1stWon",
                          "l_2ndWon","l_SvGms","l_bpSaved","l_bpFaced"), 
              selected = "minutes"))
renderPlot({
ggplot(atp,
       aes(x = atp[,input$x_axis][[1]],
           y = atp[,input$y_axis][[1]])) +
    geom_point(col = "blue") +
    geom_smooth(method = "lm", se = FALSE) +
    theme_bw() +
    xlab(input$x_axis) +
    ylab(input$y_axis)
})
```

# Embedded Application
```{r tabsets, echo=FALSE}
shiny::shinyApp(ui, server, options = list(height = 1600))
```

